Stellar Radiation is Critical for Regulating Star Formation and Driving Outflows in Low Mass Dwarf Galaxies

### 2018-08-24 ms.tex
Update on Overleaf.
added a grant number

\author{Mordecai-Mark Mac Low}
%mm [I've been asked to use my Columbia affiliation when working with CU students]
\affil{Department of Astrophysics, American Museum of Natural History, New York, NY 10024}
\affiliation{Department of Astronomy, Columbia University, New York, NY, 10027, USA}

@@

\begin{abstract}
%
%
% NOTE: Overleaf includes abstract words in word count in text... subtract
%       this from the total.
%
%
Effective stellar feedback is used in models of galaxy formation to regulate star formation and drive realistic galaxy evolution. Models typically include energy injection from supernovae as the dominant form of stellar feedback, often in some form of sub-grid recipe. However, it has been recently suggested that pre-SN feedback, from either stellar winds or stellar radiation, is necessary in high-resolution simulations of galaxy evolution to properly regulate star formation and properties of the interstellar medium (ISM). Following stellar winds and ionizing radiation is computationally challenging, so many prescriptions model this feedback approximately, accounting for the local destruction of dense gas clouds around newly formed stars in lieu of a full radiative transfer calculation. In this work we use high resolution simulations (1.8~pc) of an isolated dwarf galaxy with detailed stellar feedback tracked on a star-by-star basis. By following stellar ionizing radiation with an adaptive ray-tracing radiative transfer method, we test its importance in regulating star formation and driving outflows in this galaxy. We find that including ionizing radiation reduces the star formation rate by over a factor of 5, and is necessary in creating the ISM conditions needed for supernovae to drive significant outflows. In addition, we find that a localized approximation for radiation feedback is sufficient to regulate the star formation rate on short timescales, but is unable to drive the same outflows as in our fiducial model. Short and long range radiation effects are both important in driving the evolution of our modeled galaxy.
\end{abstract}

@@

Our fiducial simulation follows photoionizing radiation using the adaptive ray-tracing radiative transfer method of \cite{WiseAbel2011}. This method places and evolves 48 rays on a HEALPix grid around each emitting source star for both HI and HeI ionizing radiation. Rays are adaptively split as they propagate away from their source to increase the angular resolution such that the solid angle of the ray remains smaller than 1/3 of the cell area. Rays begin on HEALPix refinement level 2, with a maximum refinement level of 13. We include radiation pressure on hydrogen, but do not investigate its importance in this work \citep[see][]{WiseAbel2012}.

We additionally present a simulation run without any stellar ionizing radiation (``noRT'') and a second simulation that includes ionizing radiation, but deletes all photons that travel more than 20~pc from their source (``shortrad''). This second simulation tests the relative importance of short-range vs. long-range effects of stellar ionizing radiation as a form of feedback. Although this is still more accurate than an approximate method, this is meant to function similarly to methods that include only the local effects of stellar radiation feedback. Each of our three simulations is identical up until the formation of the first star particles.\footnote{We note that the stars from the very first star formation event in each run (21 star particles with a total mass of 114 M$_{\odot}$) are the same. Of these, one particle emits ionizing radiation ($M_* > 8 $~M$_{\odot}$).}

@@

Looking again at the first $\sim$100~Myr of simulation time, the effects of ionizing radiation beyond our 20~pc cutoff radius are not significant. However, these two simulations begin to diverge after this point. The shortrad simulation has continual, steady star formation for the entire simulation time, while the star formation rate in our fiducial run is bursty, with periods of active, low-level star formation interspersed with periods of no star formation. The driver of these differences is clear in the right hand panel. Our fiducial run loses a factor of $\sim 2$ more total gas and HI mass (black and blue lines) as compared to the shortrad simulation. Clearly, galactic winds and outflows are much more effective in with full accounting of stellar ionizing radiation feedback.

@@

\section{Discussion} \label{sec:discussion}
Accounting for feedback from stellar radiation plays a significant role in determining the ability for SN energy to couple to the ISM and therefore drive outflows. We believe this work is novel in examining the
importance of localized ionization vs.\ ionization from a diffuse radiation field far from a single star. Modeling only local stellar ionizing feedback is insufficient to describe the long-term evolution of an isolated dwarf galaxy. To explore the cause of this difference we compare in Figure~\ref{fig:panel1} and~Figure~\ref{fig:panel2} the gas number density (left), temperature (middle), and hydrogen ionization fraction (right) in edge-on slices in each of our simulations at two different times.\footnote{See https://youtu.be/r1nEW6Yu3xI for a movie of this comparison.}

\begin{figure*}
\centering
\includegraphics[width=0.99\linewidth]{DD0136_fiducial_shortrad_nort}
\caption{Edge-on slices of each simulation showing gas number density (left), temperature (middle), and hydrogen ionization fraction (right) 17~Myr after the formation of the first star in each run. Each panel is 4 kpc x 4 kpc.}
\label{fig:panel1}
\end{figure*}

@@

@@

\acknowledgments
A.E. is funded by the NSF Graduate Research Fellowship DGE 16-44869. G.L.B. is funded by NSF AST-1312888, NASA NNX15AB20G, and NSF AST-1615955. M.-M.M.L. was partly funded by NASA  grant NNX14AP27G. We gratefully recognize computational resources provided by NSF XSEDE through grant number TGMCA99S024, the NASA High-End Computing Program through the NASA Advanced Supercomputing Division at Ames Research Center, Columbia University, and the Flatiron Institute. This work made significant use of many open source software packages, including \textsc{yt}, \textsc{Enzo}, \textsc{Grackle}, \textsc{Python}, \textsc{IPython}, \textsc{NumPy}, \textsc{SciPy}, \textsc{Matplotlib}, \textsc{HDF5}, \textsc{h5py}, \textsc{Astropy}, \textsc{Cloudy} and \textsc{deepdish}. These are products of collaborative effort by many independent developers from numerous institutions around the world. Their commitment to open science has helped make this work possible.

>>>>

\author{Mordecai-Mark Mac Low}
\affil{Department of Astrophysics, American Museum of Natural History, New York, NY 10024}
\affiliation{Department of Astronomy, Columbia University, New York, NY, 10027, USA}

@@

\begin{abstract}
%
%
% NOTE: Overleaf includes abstract words in word count in text... subtract
%       this from the total.
%
%
Effective stellar feedback is used in models of galaxy formation to regulate star formation and drive realistic galaxy evolution. Models typically include energy injection from supernovae as the dominant form of stellar feedback, often in some form of sub-grid recipe. However, it has been recently suggested that pre-SN feedback (stellar winds or radiation) is necessary in high-resolution simulations of galaxy evolution to properly regulate star formation and properties of the interstellar medium (ISM). However, following these processes is computationally challenging, 
%Following stellar winds and ionizing radiation is computationally challenging, 
so many prescriptions model this feedback approximately, accounting for the local destruction of dense gas clouds around newly formed stars in lieu of a full radiative transfer calculation. In this work we examine high resolution simulations (1.8~pc) of an isolated dwarf galaxy with detailed stellar feedback tracked on a star-by-star basis. By following stellar ionizing radiation with an adaptive ray-tracing radiative transfer method, we test its importance in regulating star formation and driving outflows in this galaxy. We find that including ionizing radiation reduces the star formation rate (SFR) by over a factor of 5, and is necessary in creating the ISM conditions needed for supernovae to drive significant outflows. We find that a localized approximation for radiation feedback is sufficient to regulate the SFR on short timescales, but is unable to drive significant outflows. Short and long range radiation effects are both important in driving the evolution of our \textbf{low-metallicity, low-mass dwarf galaxy. Generalization of these results to more massive galaxies would be a valuable avenue of future research.}
\end{abstract}

@@

Our fiducial simulation follows photoionizing radiation using the adaptive ray-tracing radiative transfer method of \cite{WiseAbel2011}. This method places and evolves 48 rays on a HEALPix grid around each emitting source star for both HI and HeI ionizing radiation. Rays are adaptively split as they propagate away from their source to increase the angular resolution such that the solid angle of the ray remains smaller than 1/3 of the cell area. Rays begin on HEALPix refinement level 2, with a maximum refinement level of 13. We include radiation pressure on hydrogen, but do not investigate its importance in this work \citep[see][ and references therein]{Krumholz2018}.

We additionally present a simulation run without any stellar ionizing radiation (``noRT'') and a second simulation that includes ionizing radiation, but deletes all photons that travel more than 20~pc from their source (``shortrad''). This second simulation tests the relative importance of short-range vs. long-range effects of stellar ionizing radiation as a form of feedback. Although this is still more accurate than an approximate method, this is meant to function similarly to methods that include only the local effects of stellar radiation feedback. Each of our three simulations is identical up until the formation of the first star particles.\footnote{We note that the stars from the very first star formation event in each run (21 star particles with a total mass of 114 M$_{\odot}$) are the same. Of these, one particle emits ionizing radiation ($M_* > 8 $~M$_{\odot}$). \textbf{The sampled stellar masses differ across runs after this point. Stochastic effects from differences in IMF sampling may be important in the regime of low mass, low SFR dwarf galaxies. Multiple re-simulations of identical initial conditions may result in significant scatter in their final properties \citep{Keller2018}. We find our main conclusions are likely to be insensitive to these stochastic variations, as we find that the total radiation output across multiple re-samplings of a fixed star formation history is small after the IMF is fully sampled ($<50$~Myr).}}

\textbf{As shown in Paper I, the maximum densities reached in these simulations is below 10$^{3}$~cm$^{-3}$. Our inability to resolve the high densities in star forming regions ($\sim 10^5$~cm$^{-3}$) means that we likely underestimate initial photon absorption and overestimate the initial long-range effects of newly formed stars. However, in the Milky Way, newly formed O stars have been observed to spend no more than  10 -- 20\% of their main sequence lifetimes embedded in ultracompact HII regions within dense molecular clouds \citep{WoodChurchwell1989}. This short dispersal timescale is confirmed by high resolution simulations of massive star formation \citep[e.g.][]{Peters2010,Dale2014,Kim2018}. Therefore, we do not consider our neglect of this phase to be likely to lead to substantial errors at the galactic scale.}

@@

Looking again at the first $\sim$100~Myr of simulation time, the effects of ionizing radiation beyond our 20~pc cutoff radius are not significant. However, these two simulations begin to diverge after this point. The shortrad simulation has continual, steady star formation for the entire simulation time, while the star formation rate in our fiducial run is bursty, with periods of active, low-level star formation interspersed with periods of no star formation. The driver of these differences is clear in the right hand panel. Our fiducial run loses a factor of $\sim 2$ more total gas and HI mass (orange and blue lines) as compared to the shortrad simulation. Clearly, galactic winds and outflows are much more effective in with full accounting of stellar ionizing radiation feedback.

@@

\section{Discussion} \label{sec:discussion}
Accounting for feedback from stellar radiation plays a significant role in determining the ability for SN energy to couple to the ISM and therefore drive outflows. We believe this work is novel in examining the
importance of localized ionization vs.\ ionization from a diffuse radiation field far from a single star. Modeling only local stellar ionizing feedback is insufficient to describe the long-term evolution of an isolated dwarf galaxy. To explore the cause of this difference we compare in Figure~\ref{fig:panel1} and~Figure~\ref{fig:panel2} the gas number density (left), temperature (middle), and hydrogen ionization fraction (right) in edge-on slices in each of our simulations at two different times.

\begin{figure*}
\centering
\includegraphics[width=0.99\linewidth]{DD0136_fiducial_shortrad_nort}
\caption{Edge-on slices of each simulation showing gas number density (left), temperature (middle), and hydrogen ionization fraction (right) 17~Myr after the formation of the first star in each run. Each panel is 4 kpc x 4 kpc. \textbf{The full movie of this evolution is included for further clarification of the comparison between this figure and Fig.~\ref{fig:panel2}}}
\label{fig:panel1}
\end{figure*}

@@

\textbf{These results show that long-range ionization effects are an important consideration in models of stellar feedback. However, it warrants further study as to how much this effect can be accounted for without resorting to full radiative transfer calculations. Approximate, Stromgren-like feedback models which allow for ionization far from a single source \citep[e.x.][]{Hopkins2018} may be sufficient to capture the effects shown here. Fully localized methods, or methods which set a maximum for the ionization radius may underestimate galactic wind properties in dwarf galaxies. However, both methods are mass biased \citep[see the discussion in ][]{Hu2017}, preferentially over-ionizing dense gas that would otherwise be missed by photons leaking through channels carved through lower-density gas. These remaining uncertainties motivate a continued examination of the feedback prescriptions adopted in high resolution simulations of galaxy evolution.}

@@

\acknowledgments
A.E. is funded by the NSF Graduate Research Fellowship DGE 16-44869. G.L.B. is funded by NSF AST-1312888, NASA NNX15AB20G, and NSF AST-1615955. M.-M.M.L. was partly funded by NASA  grant NNX14AP27G and NSF grant AST18-15461. We gratefully recognize computational resources provided by NSF XSEDE through grant number TGMCA99S024, the NASA High-End Computing Program through the NASA Advanced Supercomputing Division at Ames Research Center, Columbia University, and the Flatiron Institute. This work made significant use of many open source software packages, including \textsc{yt}, \textsc{Enzo}, \textsc{Grackle}, \textsc{Python}, \textsc{IPython}, \textsc{NumPy}, \textsc{SciPy}, \textsc{Matplotlib}, \textsc{HDF5}, \textsc{h5py}, \textsc{Astropy}, \textsc{Cloudy} and \textsc{deepdish}. These are products of collaborative effort by many independent developers from numerous institutions around the world. Their commitment to open science has helped make this work possible.

\software{astropy \citep{astropy} matplotlib \citep{matplotlib}, Numpy \citep{numpy}, SciPy \citep{scipy}, IPython \citep{Ipython}, yt \citep{yt}, Enzo \citep{Enzo2014}, Grackle \citep{GrackleMethod}, CLOUDY \citep{cloudy}, HDF5}
